{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-08-09 thesis/landslide.tex
friendly dpor description (and a bit about estimation)

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\sect{\ref{sec:landslide-icb}})
	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
	\item {\tt -M}: enable Maximal State Space mode,
		which prioritizes the maximal state space to optimize for fast verification,
		abandoning all subset jobs even if they might find bugs faster
		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C})
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for P2/PSU userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
	\item {\tt -A}: support multiple abort codes during transaction failure (\sect{\ref{sec:tm-implementation}});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -S}: suppress retry aborts during transaction failure (\sect{\ref{sec:tm-implementation}})
	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\sect{\ref{sec:tm-implementation}})
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
		% TODO: put a section reference
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

@@

In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
For example, consider the execution sequence
in Figure~\ref{fig:sleepsets}(a).
After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
Figure~\ref{fig:sleepsets}(b).
However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
so from this branch,
it next tags $I_{c'1}$, causing
Figure~\ref{fig:sleepsets}(c).
to be tested next, which is regrettably equivalent to the original interleaving
($t_a$ and $t_b$ having been reordered all told, which are independent).
To recover our forgotten original purpose,
which was already fulfilled by testing the second interleaving,
DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
for an already-explored interleaving beginning with the target thread.
If such exists,
%then because of the depth-first search ordering,
the new interleaving is guaranteed to be equivalent to one already checked,
and can be skipped.

@@

Acknowledging that quickly gaining an intuitive understanding for an algorithm
is basically impossible from just paragraphs of text,
I conclude this section with an example of a single iteration of DPOR
on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.

>>>>

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\sect{\ref{sec:landslide-icb}})
	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
	\item {\tt -M}: enable Maximal State Space mode,
		which prioritizes the maximal state space to optimize for fast verification,
		abandoning all subset jobs even if they might find bugs faster
		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C}).
		According to \sect{\ref{sec:quicksand-soundness}}'s soundness proofs,
		this is equivalent to {\tt -0}
		(and according to my experience, way faster as well).
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for P2/PSU userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
	\item {\tt -A}: support multiple abort codes during transaction failure (\sect{\ref{sec:tm-implementation}});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -S}: suppress retry aborts during transaction failure (\sect{\ref{sec:tm-implementation}})
	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\sect{\ref{sec:tm-implementation}})
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
		% TODO: put a section reference
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

Although a superhuman reader may quickly reach intuitive understanding
of complex algorithms from dense prose and mathematical notation alone,
I also present here a friendly example of using DPOR on Figure~\ref{fig:concurrency-bug}'s example program,
whose original state space is shown in Figure~\ref{fig:tree}.
%Acknowledging that quickly gaining an intuitive understanding for an algorithm
%is basically impossible from just paragraphs of text,
%I conclude this section with an example of a single iteration of DPOR
%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
First, let us consider a single iteration of DPOR, applied after executing the first branch.
The result is shown in Figure~\ref{fig:dpor-example-0}.

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=\textwidth]{dpor-example-0.pdf}
	\end{center}
	\caption{Result of a single iteration of DPOR.}
	\label{fig:dpor-example-0}
\end{figure}

In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
among the threads' 4 accesses to {\tt x}.
For each such pair, it ``marks'' an alternate interleaving,
which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
These marked interleavings form a work-queue which defines the exploration.
DPOR consumes from this set in depth-first order
(note the reversed order of $\bigstar$1 and $\bigstar$2)
to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
Note also that in $\bigstar$2,
the reordered {\tt tmp2 <- x} is not directly part of the memory conflict which marked it,
but it must be executed first to reach the conflicting {\tt x <- tmp2}.

Now, let us run multiple iterations of DPOR
to advance through the first half of the full state space shown in Figure~\ref{fig:tree}(b).
Figure~\ref{fig:dpor-example-1} shows the outcome
(with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated).

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=\textwidth]{dpor-example-1.pdf}
	\end{center}
	\caption{Result of 3 DPOR iterations, pruning 7 redundant branches.}
	\label{fig:dpor-example-1}
\end{figure}

After marking $\bigstar$1 and $\bigstar$3 from Figure~\ref{fig:dpor-example-0}'s interleaving, now labeled A,
DPOR advances to interleaving B, preferring to schedule the second thread before switching back to the first
to ensure the memory access is properly reordered.
From there, it identifies a new memory conflict, marks $\bigstar$2, and advances to C,
where it finds no memory conflicts that would mark anything not already marked and/or explored
(memory conflicts that were already reordered in old branches are not highlighted with arrows).
From C, $\bigstar$3 alone remains in the work-queue,
so DPOR advances to the second (symmetric) half of the state space,
skipping (thereby pruning) branches D through J.

To see why branches D through J need not be tested, consider that each thread's {\tt tmpN++} is a thread-local event,
participating in no memory conflicts,
and hence any two interleavings differing only by reordering {\tt tmpN++}s must be equivalent.
The dashed blue arrows denote such equivalences;
note the two disjoint equivalence classes \{B,E,G,I\} and \{C,D,F,H,J\},
distinguished by the order of the two final {\tt x <- tmpN}s.
Note also that
%(as shown in Figure~\ref{fig:tree}(b)),
although B and C also have the same outcome ({\tt x==1}),
this depends on the {\em values} written to memory rather than {\em addresses}
(and would change if one thread's {\tt tmpN++} were a {\tt tmpN+=2}, for example),
which DPOR does not consider.
Recent work \cite{mcr} has extended DPOR to find such value-based equivalences,
although is beyond this explanation's scope.

Finally, let us consider the final result after DPOR runs out of remaining unexplored marked branches,
shown in Figure~\ref{fig:dpor-example-2}.

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=\textwidth]{dpor-example-2.pdf}
	\end{center}
	\caption{DPOR's termination state, having reduced 20 interleavings to 6.}
	\label{fig:dpor-example-2}
\end{figure}

Ultimately, the second half of the state space is pruned symmetrically.
In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$;
in this case, the original state space's size was ${3+3 \choose 3} = 20$.
DPOR's reduction is characterized by replacing $N$ and $M$ with the number of {\em conflicting} events only;
in this case, ignoring all {\tt tmpN++} reorderings and testing only ${2+2 \choose 2} = 6$ branches.

However, as the legend spoils, this subset of the state space still contains redundant interleavings,
arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
Visual inspection shows that $\bigstar$5, while locally justified in trying to reorder {\tt tmp1 <- x} before {\tt x <- tmp2},
effectively serves % oops
only to reorder it with {\tt tmp2 <- x}
relative to the first symmetric subtree.
Landslide identifies this equivalence by checking,
%in this case during the iteration of DPOR after interleaving K,
in this case after K,
that if the first event to be reordered (here, {\tt tmp <- x})
has already been tested in an equivalent reordering around any number of preceding events (here, {\tt tmp2 <- x} and {\tt tmp2++}),
then the newly marked subtree is safe to prune.
The next section describes the implementation of this check in general.

@@

% TODO: condense this by removing the 'new' example and just continuing with the previous section's
In the presence of non-conflicting transitions as well as conflicting ones,
DPOR's approach as described so far can still end up testing equivalent interleavings.
For example, consider the execution sequence
in Figure~\ref{fig:sleepsets}(a).
After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
Figure~\ref{fig:sleepsets}(b).
However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
so from this branch,
it next tags $I_{c'1}$, causing
Figure~\ref{fig:sleepsets}(c).
to be tested next, which is regrettably equivalent to the original interleaving
($t_a$ and $t_b$ having been reordered all told, which are independent).
To recover our forgotten original purpose,
which was already fulfilled by testing the second interleaving,
DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
for an already-explored interleaving beginning with the target thread.
If such exists,
%then because of the depth-first search ordering,
the new interleaving is guaranteed to be equivalent to one already checked,
and can be skipped.

@@

For both the user's convenience and for Quicksand's prioritization algorithm (\sect{\ref{sec:quicksand-id}}),
Landslide attempts to guess how big partially-explored state spaces will ultimately end up being upon completion.
Because the backtracking implementation uses checkpointing
rather than replaying similar interleavings' shared execution prefixes from the beginning \sect{\ref{sec:landslide-timetravel}},
the total number of interleavings (i.e., leaf nodes in the execution tree)
must be estimated separately from the total runtime (i.e., sum of all edge weights in the tree).

% TODO: once dpor example tree is made
% switch this to that
As a concrete example, consider the state space of Figure~\ref{fig:tree}(b) (untouched by DPOR),
and suppose each transition to the next preemption point takes 1 second to execute.
While the first branch executes in 5 seconds,
the second branch, sharing the first transition as a common prefix, takes 4 seconds,
and the one after that only 2;
the state space being ultimately completed in 33 seconds.
Even with perfect hindsight,
na\"ively multiplying the total interleavings (10)
by the total execution time per branch (5) would double-count common prefixes
and grossly overestimate (50) the total runtime.

Hence, Landslide uses two differently suitable algorithms for each of size and runtime estimation:
the Weighted Backtrack Estimator (WBE) and the Recursive Estimator (RE), respectively,
first introduced in \cite{estimating-search-tree-size} and later adapted to DPOR by \cite{estimation}.
In principle, both calculate the current progress as a proportion of the expected total
by counting how many branches DPOR has marked for future exploration (\sect{\ref{sec:landslide-explore}})
and assuming the sizes of their resulting subtrees are predicted by the known sizes of similar already-explored subtrees.
In practice, the calculation strategy differs between the two approaches,
which can occasionally result in drastically differing outputs (\sect{\ref{sec:tm-verif}}).
I summarize their respective implementations here.
% TODO more specific section reference
Chapter~\ref{chap:warpzone} will discuss their limitations and some opportunities for future improvement.

\subsubsection{Size (Weighted Backtrack) estimation}

\subsubsection{Runtime (Recursive) estimation}

### 2018-08-09 thesis/overview.tex
revise tm, cleanup todos in bgnd/landslide, fix cross-chapter section references, etc -- sending tm chapter off to committee here
friendly dpor description (and a bit about estimation)

\begin{figure}[h]
	\input{fig-htm-example}
	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
		If the transaction in the top branch aborts,
		whether from a memory conflict or random system interrupt,
		%from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin},
		{\tt status} will be assigned an error code indicating the abort reason,
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

@@

\begin{figure}[p]
	\begin{tabular}{c}
		\includegraphics[width=\textwidth]{statespace-list.pdf}
		\\
		(a) Interleavings visualized individually, as a list.
		\\
		\\
		\includegraphics[width=\textwidth]{statespace-tree.pdf}
		\\
		(b) Interleavings (same order as in (a)), with common prefixes \\
		combined as ``preemption points'', forming a tree.
	\end{tabular}
	\caption{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
	Thread 1 is represented by blue ovals, thread 2 by orange squares, and time flows from top to bottom.
	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
	the full state space is twice the size, and the other half is symmetric to the one shown.}
	\label{fig:tree}
\end{figure}

@@

Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.

>>>>

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{l}
		\texttt{\ctype{void} \call{count}() \{} \\
		\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
		\texttt{~~~~~~~~\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\call{\_xend}();} \\
		\texttt{~~~~~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~~~~~\call{mutex\_lock}(\&m);} \\
		\texttt{~~~~~~~~~~~~x++;} \\
		\texttt{~~~~~~~~~~~~\call{mutex\_unlock}(\&m);} \\
		\texttt{~~~~~~~~\}} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
		If the transaction in the top branch aborts,
		whether from a memory conflict or random system interrupt,
		%from the programmer's intention,
		execution will revert to the return of {\tt \_xbegin},
		{\tt status} will be assigned an error code indicating the abort reason,
		and control will drop into the {\tt else} branch.
		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
	\label{fig:htm-example}
\end{figure}

@@

\begin{figure}[p]
	\begin{tabular}{c}
		\includegraphics[width=\textwidth]{statespace-list.pdf}
		\\
		(a) Interleavings visualized individually, as a list.
		\\
		\\
		\includegraphics[width=\textwidth]{statespace-tree.pdf}
		\\
		(b) Interleavings (same order as in (a)), with common prefixes \\
		combined as ``preemption points'', forming a tree.
	\end{tabular}
	\caption{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
	Thread 1 is represented by purple ovals, thread 2 by yellow squares, and time flows from top to bottom.
	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
	the full state space is twice the size, and the other half is symmetric to the one shown.}
	\label{fig:tree}
\end{figure}

@@

Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the most popular algorithm for mitigating the exponential explosion that arises as program size increases.
